---
title: "individual1.Rmd"
author: "Nathan Marquardt"
date: "3/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part one: Cleaning Data

Although the three Excel sheets are loaded into R, they are not useful before
some cleaning. Manipulate the three imported data frames, so that in the cleaned dataset:

1.The player names and opponent names are factors.

2.All offensive and defensive statistics are numerical. The `Weight` column in biography is also numeric. (Optional: Read `?across` and learn to mutate multiple columns simultaneously using the `across` function in `dplyr`)

3.Change the `Height` column in biography into numeric. (Hint: Use a unit that works well)

  Answer for the Part 1 Questions:
  ```{r}
  library(readxl) 
library(dplyr)
library(tidyr)
defense <- read_excel('/Users/nathanmarquardt/Desktop/ds202_lab4/cyclonesFootball2020.xlsx', sheet='Defensive')
offense <- read_excel('/Users/nathanmarquardt/Desktop/ds202_lab4/cyclonesFootball2020.xlsx', sheet='Offensive')
biography <- read_excel('/Users/nathanmarquardt/Desktop/ds202_lab4/cyclonesFootball2020.xlsx', sheet='Biography')
str(defense)
?across

#Part 1 Question 1
isuOffPlayers <- factor(offense$Name)
isuOffOpponents <- factor(offense$Opponent_Opponent)

isuDefPlayers <- factor(defense$Name)
isuDefOpponents <- factor(defense$Opponent_Opponent)


#Part 1 Question 2
defClean <- defense %>%
                mutate(across(Tackles_Solo:Pass_PB, as.numeric))


offClean <- offense %>%
  mutate(across(Receiving_REC:Passing_INT, na.rm = TRUE, as.numeric))


#Part 1 Question 3
bioClean <- biography %>%
              separate(Height, c('feet', 'inches'), "-", convert = TRUE) %>%
                mutate(Total_Inches = (12 * feet) + inches)
                
str(offClean)
str(defClean)
str(bioClean)
```  

## Part two: Tidying

For the following questions, work on the cleaned data frames `defClean`, `offClean`, and `bioClean`.

1.Reformat the `offClean` data frame into a tidy format using `pivot_longer`. The type of statistic (Receiving_REC for the number of receivings, Receiving_YDS for receiving yards) is added as a new key column named `stat`.

  Answer:
  
2.Show on the distributions of the offensive statistics. The distribution of each offensive statistic should be shown in a separate facet. Explain the patterns you see.

  Answer:
  
3.Did ISU have better offensive against Oregon or Oklahoma? Answer this question by creating a scatterplot with the x- and y-axes being the receiving yards of the ISU players against these two opponents. A large number of receiving yards indicates good offense.

  Answer:
  
Now turn our attention to the player biography.

4.Separate the `Hometown` column into two columns, namely, the city and the state where the player comes from. Print the fiers twe rows of these two columns. (Hint: look at the `sep=` argument of `separate`)

  Answer:
  
5.How many players are there from each state?

  Answer:
  

## Part three: Joining data frames

1.Find which states do the ISU players (both defensive and offensive) come from. Create an appropriate numerical or graphical summary.

  Answer:
  
2.How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?

  Answer:
  
3.Which ISU player(s) made the biggest progress over the last year? Compare the player-level statistics (i.e. averages and totals for each player) using both the 2020 and the 2019 data here. Choose for yourself the player statistics to investigate.

  Answer:
